---
title: "DATA608 Assignment 7"
author: "Steven Gonzalez"
output: html_document
---

## Prompt
The US Geological Survey publishes a list of Strategic Minerals (https://www.usgs.gov/news/national-news-release/us-geological-survey-releases-2022-list-critical-minerals ). Having a secure supply of these minerals is essential to our security and economic prosperity. However, many of these minerals are sourced from outside of the US. This assignment is to develop a reference catalog of the source or sources of each of these minerals and a judgement on the reliability of each source under stressed circumstance (e.g. war, economic crisis, etc.).

## R Packages
```{r message = FALSE, warning = FALSE}
library(tidyverse)
library(httr)
library(jsonlite)
library(dplyr)
library(ggplot2)
library(reshape2)
library(usmap)
library(tinytex)
library(paletteer)
library(gapminder)
```

## Data Import
```{r message = FALSE, warning = FALSE}
url <- "https://raw.githubusercontent.com/Stevee-G/Data608/refs/heads/main/Assignment7/min_list.csv"
url1 <- "https://raw.githubusercontent.com/Stevee-G/Data608/refs/heads/main/Assignment7/ofr20051294/deposit.csv"
url2 <- "https://raw.githubusercontent.com/Stevee-G/Data608/refs/heads/main/Assignment7/us_allies.csv"
url3 <- "https://raw.githubusercontent.com/Stevee-G/Data608/refs/heads/main/Assignment7/us_rivals.csv"

min_list <- read_csv(url)
minerals <- read_csv(url1)
allies <- read_csv(url2)
competitors <- read_csv(url3)

glimpse(min_list)
glimpse(minerals)
glimpse(allies)
glimpse(competitors)
```

## Data Wrangling
```{r message = FALSE, warning = FALSE}
min_trim <- minerals %>% 
  select(commodity, country,
         latitude, longitude) %>% 
  arrange(commodity, country)
glimpse(min_trim)

countries <- bind_rows(allies, competitors)
countries$reliability <- case_when(
  countries$Status == 'Ally' ~ 'Ally',
  countries$Status == 'Former enemy' ~ 'Neutral',
  countries$Status == 'Enemy' ~ 'Competitor'
  )
countries <- countries %>% 
  select(Country, reliability)
glimpse(countries)

all <- right_join(countries,
                  inner_join(min_trim,
                             min_list,join_by(commodity == Mineral),
                             relationship = 'many-to-one'),
                  join_by(Country == country),
                  relationship = 'one-to-many') %>% 
  arrange(commodity, Country)
all$reliability <- all$reliability %>% 
  replace_na('Neutral')
all$line_score <- case_when(
  all$reliability == 'Ally' ~ 1,
  all$reliability == 'Neutral' ~ 0,
  all$reliability == 'Competitor' ~ -1
)
glimpse(all)

availability <- all %>% 
  group_by(commodity) %>% 
  summarize(
    '# countries' = n(),
    access = sum(line_score)
  ) %>% 
  ungroup()
glimpse(availability)
```

## Data Visualizations
```{r message = FALSE, warning = FALSE}
ggplot(availability, aes(reorder(commodity, `# countries`),
                         `# countries`), fill = `# countries`) + 
  geom_bar(stat = "identity") +
  scale_fill_paletteer_c('ggthemes::Blue-Green Sequential') +
  labs(title = "Barplot: Number of Countries Mining Each Mineral",
       x = "") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  theme(plot.title = element_text(hjust = 0.5)) +
  coord_flip()

ggplot(availability, aes(reorder(commodity, access),
                         access), fill = access) + 
  geom_bar(stat = "identity") +
  scale_fill_paletteer_c('ggthemes::Blue-Green Sequential') +
  labs(title = "Barplot: Mineral Reliability",
       x = "") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  theme(plot.title = element_text(hjust = 0.5)) +
  coord_flip()
```
```{r message = FALSE, warning = FALSE}

```

## Sources
Mineral List:

https://www.usgs.gov/news/national-news-release/us-geological-survey-releases-2022-list-critical-minerals

Mineral Data:

https://mrdata.usgs.gov/major-deposits/

Ally Data:

https://worldpopulationreview.com/country-rankings/united-states-allies

Competitor Data:

https://worldpopulationreview.com/country-rankings/us-enemy-countries
