---
title: "DATA608 Assignment 7"
author: "Steven Gonzalez"
output: html_document
---

## Prompt
The US Geological Survey publishes a list of Strategic Minerals (https://www.usgs.gov/news/national-news-release/us-geological-survey-releases-2022-list-critical-minerals ). Having a secure supply of these minerals is essential to our security and economic prosperity. However many of these minerals are sourced from outside of the US. This assignment is to develop a reference catalog of the source or sources of each of these minerals and a judgement on the reliability of each source under stressed circumstance (e.g. war, economic crisis, etc.).

## R Packages
```{r message = FALSE, warning = FALSE}
library(tidyverse)
library(httr)
library(jsonlite)
library(dplyr)
library(ggplot2)
library(reshape2)
library(usmap)
library(tinytex)
library(paletteer)
```

## Data Import
```{r message = FALSE, warning = FALSE}
url <- "https://raw.githubusercontent.com/Stevee-G/Data608/refs/heads/main/Assignment7/ofr20051294/deposit.csv"
url1 <- "https://raw.githubusercontent.com/Stevee-G/Data608/refs/heads/main/Assignment7/us_allies.csv"
url2 <- "https://raw.githubusercontent.com/Stevee-G/Data608/refs/heads/main/Assignment7/us_rivals.csv"

minerals <- read_csv(url)
allies <- read_csv(url1)
competitors <- read_csv(url2)

glimpse(minerals)
glimpse(allies)
glimpse(competitors)
```

## Data Wrangling
```{r message = FALSE, warning = FALSE}
min_trim <- minerals %>% 
  select(commodity, country,
         latitude, longitude) %>% 
  arrange(commodity, country)
glimpse(min_trim)


```

## Data Visualizations
```{r message = FALSE, warning = FALSE}

```

## Sources
Mineral Data:

https://mrdata.usgs.gov/major-deposits/

Ally Data:

https://worldpopulationreview.com/country-rankings/united-states-allies

Competitor Data:

https://worldpopulationreview.com/country-rankings/us-enemy-countries
