---
title: "DATA608 Assignment 3"
author: "Steven Gonzalez"
output: html_document
---

## Prompt
The CDC publishes firearm mortality for each State per 100,000 persons (https://www.cdc.gov/nchs/state-stats/deaths/firearms.html). Each State' firearm control laws can be categorized as very strict to very lax. The purpose of this Story is to answer the question, " Do stricter firearm control laws help reduce firearm mortality?"
For this assignment you will need to:
Access the firearm mortality data from the CDC using an available API (https://open.cdc.gov/apis.html)
Create a 5 point Likert scale categorizing gun control laws from most lax to strictest and assign each state to the most appropriate Likert bin.
Determine whether stricter gun control laws result in reduced gun violence deaths
Present your story using  heat maps

## R Packages
```{r message = FALSE, warning = FALSE}
library(tidyverse)
library(httr)
library(jsonlite)
library(dplyr)
```

## Data Import
```{r message = FALSE, warning = FALSE}
url <- "https://data.cdc.gov/api/v3/views/fpsi-y8tj/query.json"
key <- "MCuyIrLZPnjyG1E7PNCJgbJXZ"
url1 <- "https://raw.githubusercontent.com/Stevee-G/Data608/refs/heads/main/Assignment3/firearm_laws.csv"
url2 <- "https://raw.githubusercontent.com/Stevee-G/Data608/refs/heads/main/Assignment3/law_rankings.csv"

response <- GET(
  url = url,
  add_headers(`X-App-Token` = key)
)
status_code(response)
stats <- fromJSON(content(response, "text"))

laws <- read_csv(url1)

ranks <- read_csv(url2)

glimpse(stats)
glimpse(laws)
glimpse(ranks)
```

## Data Wrangling
```{r message = FALSE, warning = FALSE}
stats_trim <- stats %>% 
  filter(period == "2024" & (intent == "FA_Deaths" |
                              intent == "FA_Homicide" |
                              intent == "FA_Suicide")) %>% 
  mutate(count_sup = as.integer(count_sup),
         rate = as.double(rate)) %>% 
  select(name, intent, count_sup, rate)
glimpse(stats_trim)

laws_trim <- laws %>% 
  arrange(desc(law_id)) %>% 
  mutate(bridge = paste(law_class,
                        law_class_subtype,
                        handguns_or_long_guns)) %>%
  distinct(paste(state, bridge), .keep_all = TRUE) %>% 
  arrange(law_id) %>% 
  select(state, bridge)
glimpse(laws_trim)

#state_rank <- 


cpi_trim <- cpi_unnested %>%
  mutate(date = as.Date(yearmonth(paste(
    year, periodName, sep = " ")), format = "%Y %b")) %>%
  rename(inflation_rate = "12") %>%
  select(date, inflation_rate) %>%
  arrange(date)

cpi_trim$inflation_rate <- as.numeric(cpi_trim$inflation_rate)
glimpse(cpi_trim)

wg_dated <- wg_data %>% 
  mutate(date = as.Date(Monthyear, format = "%m/%d/%Y")) %>%
  rename(wage_growth = Overall) %>% 
  select(date, wage_growth)

wg_trim <- wg_dated %>% 
  filter(date >= "2004-01-01" & date <= "2023-12-01")

wg_trim$wage_growth <- as.numeric(wg_trim$wage_growth)
glimpse(wg_trim)

combined_data <- inner_join(cpi_trim, wg_trim, by = join_by(date))

data <- combined_data %>%
  rename("Inflation Rate" = inflation_rate) %>%
  rename("Wage Growth" = wage_growth) %>%
  pivot_longer(
    cols = c("Inflation Rate", "Wage Growth"),
    names_to = "measure",
    values_to = "value"
  )
glimpse(data)
```

## Data Visualizations
```{r message = FALSE, warning = FALSE}
summary(data$value)

describe(data$value)

description <- describeBy(data$value, group = data$measure, mat = TRUE, skew = FALSE)
names(description)[2] <- "measure"
description$Var <- description$sd^2
description$contrast <- (description$mean - mean(description$mean))
description

ggplot(data = data, aes(x = date, y = value, color = measure, shape = measure)) +
  geom_line() +
  geom_point() +
  labs(title = "Inflation Rate and Wage Growth vs Time",
       x = "Date", y = "% Change", color = NULL, shape = NULL) +
  theme(plot.title = element_text(hjust = 0.5))

ggplot(data = data, aes(x = measure, y = value, fill = measure)) +
  geom_boxplot() +
  labs(title = "Average Change by Inflation Rate and Wage Growth",
       x = "Measure", y = "% Change", fill = NULL) +
  theme(plot.title = element_text(hjust = 0.5))
```

## Sources
CDC FA Mortality:
(https://data.cdc.gov/Injury-Violence/Mapping-Injury-Overdose-and-Violence-State/fpsi-y8tj/about_data)

Rand FA Laws:
(https://www.rand.org/research/gun-policy/law-navigator.html#minimum-age-for-possession)