---
title: "DATA608 Assignment 5"
author: "Steven Gonzalez"
output: html_document
---

## Prompt
Historic data indicates that the occurrence and intensity of cyclonic storms (Hurricanes, Typhoons and Tornadoes) increases with the increased earth temperature. For this assignment you will need to tell this story to a non-technical audience (eg: a high-school earth science class).

## R Packages
```{r message = FALSE, warning = FALSE}
library(tidyverse)
library(httr)
library(jsonlite)
library(dplyr)
library(ggplot2)
library(reshape2)
library(usmap)
library(tinytex)
```

## Data Import
```{r message = FALSE, warning = FALSE}
url <- "https://www.nhc.noaa.gov/data/hurdat/hurdat2-1851-2025-02272026.txt"
url1 <- "https://www.nhc.noaa.gov/data/hurdat/hurdat2-nepac-1949-2025-02272026.txt"
url2 <- "https://data.giss.nasa.gov/gistemp/graphs/graph_data/Global_Mean_Estimates_based_on_Land_and_Ocean_Data/graph.txt"

out_file <- "hurdat2_atlantic.txt"
download.file(url, out_file, mode = "wb")
raw <- readLines(out_file)

parsed <- list()
i <- 1
while (i <= length(raw)) {
  header <- trimws(strsplit(raw[i], ",")[[1]])
  track_id <- header[1]
  track_name <- header[2]
  rows <- as.numeric(header[3])

  body <- raw[(i + 1):(i + rows)]
  body <- strsplit(body, ",")
  body <- lapply(body, trimws)
  body <- lapply(body, function(x) x[1:20])  # keep first 20 fields
  body <- do.call(rbind, body)

  colnames(body) <- c(
    "date","time","record","status","lat","lon","max_speed","min_press",
    "34kt_ne","34kt_se","34kt_sw","34kt_nw",
    "50kt_ne","50kt_se","50kt_sw","50kt_nw",
    "64kt_ne","64kt_se","64kt_sw","64kt_nw"
  )

  body <- as_tibble(body)
  body$lat <- ifelse(grepl("N", body$lat),
                     gsub("N","", body$lat),
                     paste0("-", gsub("S","", body$lat)))
  body$lon <- ifelse(grepl("E", body$lon),
                     gsub("E","", body$lon),
                     paste0("-", gsub("W","", body$lon)))

  parsed[[length(parsed) + 1]] <-
    bind_cols(tibble(track_id = track_id, track_name = track_name), body)

  i <- i + rows + 1
}

atlantic_df <- bind_rows(parsed)
head(atlantic_df)

out_file1 <- "hurdat2_pacific.txt"
download.file(url1, out_file1, mode = "wb")
raw1 <- readLines(out_file1)

parsed1 <- list()
j <- 1
while (j <= length(raw1)) {
  header <- trimws(strsplit(raw1[j], ",")[[1]])
  track_id <- header[1]
  track_name <- header[2]
  rows <- as.numeric(header[3])

  body <- raw1[(j + 1):(j + rows)]
  body <- strsplit(body, ",")
  body <- lapply(body, trimws)
  body <- lapply(body, function(x) x[1:20])  # keep first 20 fields
  body <- do.call(rbind, body)

  colnames(body) <- c(
    "date","time","record","status","lat","lon","max_speed","min_press",
    "34kt_ne","34kt_se","34kt_sw","34kt_nw",
    "50kt_ne","50kt_se","50kt_sw","50kt_nw",
    "64kt_ne","64kt_se","64kt_sw","64kt_nw"
  )

  body <- as_tibble(body)
  body$lat <- ifelse(grepl("N", body$lat),
                     gsub("N","", body$lat),
                     paste0("-", gsub("S","", body$lat)))
  body$lon <- ifelse(grepl("E", body$lon),
                     gsub("E","", body$lon),
                     paste0("-", gsub("W","", body$lon)))

  parsed1[[length(parsed1) + 1]] <-
    bind_cols(tibble(track_id = track_id, track_name = track_name), body)

  j <- j + rows + 1
}

pacific_df <- bind_rows(parsed1)
head(pacific_df)

out_file2 <- "nasa_temps.txt"
download.file(url2, out_file2, mode = "wb")
temp_df <- readLines(out_file2)
temp_df <- read_csv(url2)
glimpse(temp_df)
```

## Data Wrangling
```{r message = FALSE, warning = FALSE}

```

## Data Visualizations
```{r message = FALSE, warning = FALSE}
plot_usmap(data = state, values = "mean", color = "black") +
  scale_fill_continuous(low = "whitesmoke",
                        high = "darkblue",
                        name = "Average Salary",
                        label = scales::comma) +
  labs(title = "Heatmap: Data Practitioner Salary Across the States") +
  theme(legend.position = "right") +
  theme(plot.title = element_text(hjust = 0.5))
```
```{r message = FALSE, warning = FALSE, fig.dim = c(5,10)}
ggplot(state_long, aes(x = value, y = reorder(state, value, FUN = max))) +
  geom_line() +
  geom_point(aes(color = variable), size = 2) +
  scale_color_brewer(palette = "Set1", direction = -1) +
  labs(title = "Dumbbell Plot: Salary Range by State",
       x = "",
       y = "") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  theme(plot.title = element_text(hjust = 0.5))
```
```{r message = FALSE, warning = FALSE}
ggplot(prof_long, aes(x = value, y = reorder(OCC_TITLE, value, FUN = max))) +
  geom_line() +
  geom_point(aes(color = variable), size = 2) +
  scale_color_brewer(palette = "Set1", direction = -1) +
  labs(title = "Dumbbell Plot: Salary Range by Profession",
       x = "",
       y = "") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  theme(plot.title = element_text(hjust = 0.5))
```
```{r message = FALSE, warning = FALSE, fig.dim = c(10,6)}
ggplot(field_long, aes(x = value, y = reorder(NAICS_TITLE, value, FUN = max))) +
  geom_line() +
  geom_point(aes(color = variable), size = 2) +
  scale_color_brewer(palette = "Set1", direction = -1) +
  labs(title = "Dumbbell Plot: Salary Range by Industry",
       x = "",
       y = "") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  theme(plot.title = element_text(hjust = 0.5))
```

## Sources
Storm Data:

https://www.nhc.noaa.gov/data/#hurdat

Climate Data:

https://science.nasa.gov/earth/explore/earth-indicators/global-temperature/
