---
title: "DATA608 Assignment 6"
author: "Steven Gonzalez"
output: html_document
---

## Prompt
The United Nations Food and Agriculture Organization publication, The State of Food Security and Nutrition in the World 2022 (https://www.fao.org/documents/card/en/c/cc0639en) might lead one to the conclusion that it's an elsewhere problem. That the people who are suffering malnutrition and starvation are "elsewhere", not in our backyard. For this assignment you will need to take a closer look here at home (the US).

## R Packages
```{r message = FALSE, warning = FALSE}
library(tidyverse)
library(httr)
library(jsonlite)
library(dplyr)
library(ggplot2)
library(reshape2)
library(usmap)
library(tinytex)
library(paletteer)
```

## Data Import
```{r message = FALSE, warning = FALSE}
url <- "https://raw.githubusercontent.com/Stevee-G/Data608/refs/heads/main/Assignment6/foodsecurity_csv_datafiles/foodsecurity-all-households-2024.csv"
url1 <- "https://raw.githubusercontent.com/Stevee-G/Data608/refs/heads/main/Assignment6/foodsecurity_csv_datafiles/foodsecurity-child-trends-2024.csv"
url2 <- "https://raw.githubusercontent.com/Stevee-G/Data608/refs/heads/main/Assignment6/foodsecurity_csv_datafiles/foodsecurity-educ-emp-dis-2024.csv"
url3 <- "https://raw.githubusercontent.com/Stevee-G/Data608/refs/heads/main/Assignment6/foodsecurity_csv_datafiles/foodsecurity-state-2024.csv"
url4 <- "https://raw.githubusercontent.com/aruljohn/us-states/refs/heads/master/states.csv"

all_hh <- read_csv(url)
ct <- read_csv(url1)
eed <- read_csv(url2)
state <- read_csv(url3)
state_abbr <- read_csv(url4)

glimpse(all_hh)
glimpse(ct)
glimpse(eed)
glimpse(state)
glimpse(state_abbr)
```

## Data Wrangling
```{r message = FALSE, warning = FALSE}
hh_comp <- all_hh %>% 
  filter(Category == "Household composition") %>% 
  group_by(Subcategory) %>% 
  summarize(
    `Food secure-percent` = round(mean(`Food secure-percent`),2),
    `Food insecure-percent` = round(mean(`Food insecure-percent`),2),
    `Low food security-percent` = round(mean(`Low food security-percent`),2),
    `Very low food security-percent` = round(mean(`Very low food security-percent`),2)
  ) %>% 
  pivot_longer(
    cols = c(`Food secure-percent`, `Food insecure-percent`,
             `Low food security-percent`, `Very low food security-percent`),
    names_to = "Food Security Type",
    values_to = "Percent"
  )
glimpse(hh_comp)

re_hh <- all_hh %>% 
  filter(Category == "Race/ethnicity of households") %>% 
  group_by(Subcategory) %>% 
  summarize(
    `Food secure-percent` = round(mean(`Food secure-percent`),2),
    `Food insecure-percent` = round(mean(`Food insecure-percent`),2),
    `Low food security-percent` = round(mean(`Low food security-percent`),2),
    `Very low food security-percent` = round(mean(`Very low food security-percent`),2)
  ) %>% 
  pivot_longer(
    cols = c(`Food secure-percent`, `Food insecure-percent`,
             `Low food security-percent`, `Very low food security-percent`),
    names_to = "Food Security Type",
    values_to = "Percent"
  )
glimpse(re_hh)

ipr <- all_hh %>% 
  filter(Category == "Household income-to-poverty ratio") %>% 
  group_by(Subcategory) %>% 
  summarize(
    `Food secure-percent` = round(mean(`Food secure-percent`),2),
    `Food insecure-percent` = round(mean(`Food insecure-percent`),2),
    `Low food security-percent` = round(mean(`Low food security-percent`),2),
    `Very low food security-percent` = round(mean(`Very low food security-percent`),2)
  ) %>% 
  pivot_longer(
    cols = c(`Food secure-percent`, `Food insecure-percent`,
             `Low food security-percent`, `Very low food security-percent`),
    names_to = "Food Security Type",
    values_to = "Percent"
  )
glimpse(ipr)

ar <- all_hh %>% 
  filter(Category == "Area of residence") %>% 
  group_by(Subcategory) %>% 
  summarize(
    `Food secure-percent` = round(mean(`Food secure-percent`),2),
    `Food insecure-percent` = round(mean(`Food insecure-percent`),2),
    `Low food security-percent` = round(mean(`Low food security-percent`),2),
    `Very low food security-percent` = round(mean(`Very low food security-percent`),2)
  ) %>% 
  pivot_longer(
    cols = c(`Food secure-percent`, `Food insecure-percent`,
             `Low food security-percent`, `Very low food security-percent`),
    names_to = "Food Security Type",
    values_to = "Percent"
  )
glimpse(ar)

cgr <- all_hh %>% 
  filter(Category == "Census geographic region") %>% 
  group_by(Subcategory) %>% 
  summarize(
    `Food secure-percent` = round(mean(`Food secure-percent`),2),
    `Food insecure-percent` = round(mean(`Food insecure-percent`),2),
    `Low food security-percent` = round(mean(`Low food security-percent`),2),
    `Very low food security-percent` = round(mean(`Very low food security-percent`),2)
  ) %>% 
  pivot_longer(
    cols = c(`Food secure-percent`, `Food insecure-percent`,
             `Low food security-percent`, `Very low food security-percent`),
    names_to = "Food Security Type",
    values_to = "Percent"
  )
glimpse(cgr)

ct_trim <- ct %>% 
  select(Category, `Food-secure households-percent`,
         `Food-insecure households-percent`,
         `Households with food-insecure children-percent`,
         `Households with very low food security among children-percent`) %>% 
  pivot_longer(
    cols = c(`Food-secure households-percent`, `Food-insecure households-percent`,
             `Households with food-insecure children-percent`,
             `Households with very low food security among children-percent`),
    names_to = "Food Security Type",
    values_to = "Percent"
  )
glimpse(ct_trim)

emp <- eed %>% 
  filter(Subcategory == 'Employment') %>% 
  select(`Sub-subcategory`, `Food insecure-percent`,
         `Very low food security-percent`) %>% 
  pivot_longer(
    cols = c(`Food insecure-percent`,`Very low food security-percent`),
    names_to = "Food Security Type",
    values_to = "Percent"
  )
glimpse(emp)

edu <- eed %>% 
  filter(Subcategory == 'Education') %>% 
  select(`Sub-subcategory`, `Food insecure-percent`,
         `Very low food security-percent`) %>% 
  pivot_longer(
    cols = c(`Food insecure-percent`,`Very low food security-percent`),
    names_to = "Food Security Type",
    values_to = "Percent"
  )
glimpse(edu)

dis <- eed %>% 
  filter(Subcategory == 'Disability status') %>% 
  select(`Sub-subcategory`, `Food insecure-percent`,
         `Very low food security-percent`) %>% 
  pivot_longer(
    cols = c(`Food insecure-percent`,`Very low food security-percent`),
    names_to = "Food Security Type",
    values_to = "Percent"
  )
glimpse(dis)

state_trim <- inner_join(state, state_abbr,
                        join_by(State == ABBREVIATION),
                        relationship = "one-to-one") %>% 
  select(NAME, `Food insecurity prevalence`,
         `Very low food security prevalence`) %>% 
  mutate(state = NAME)
glimpse(state_trim)
```

## Data Visualizations
```{r message = FALSE, warning = FALSE}
plot_usmap(data = state_trim, values = "Food insecurity prevalence", color = "black") +
  scale_fill_continuous(low = "whitesmoke",
                        high = "darkorange",
                        name = "Level of Prevalence",
                        label = scales::comma) +
  labs(title = "Heatmap: Food Insecurity Prevalence Across the States") +
  theme(legend.position = "right") +
  theme(plot.title = element_text(hjust = 0.5))

plot_usmap(data = state_trim, values = "Very low food security prevalence", color = "black") +
  scale_fill_continuous(low = "whitesmoke",
                        high = "darkred",
                        name = "Level of Prevalence",
                        label = scales::comma) +
  labs(title = "Heatmap: Very Low Food Security Prevalence Across the States") +
  theme(legend.position = "right") +
  theme(plot.title = element_text(hjust = 0.5))
```
```{r message = FALSE, warning = FALSE, fig.dim = c(10,7)}
ggplot(hh_comp, aes(x = Subcategory, y = Percent, fill = `Food Security Type`)) + 
  geom_bar(stat = "identity") +
  scale_fill_paletteer_d('ggsci::orange_material') +
  labs(title = "Barplot: Food Security by Household Composition") +
  theme_minimal() +
  theme(legend.position = "right") +
  theme(plot.title = element_text(hjust = 0.5))

ggplot(re_hh, aes(x = Subcategory, y = Percent, fill = `Food Security Type`)) + 
  geom_bar(stat = "identity") +
  scale_fill_paletteer_d('ggsci::orange_material') +
  labs(title = "Barplot: Food Security by Race/Ethnicity") +
  theme_minimal() +
  theme(legend.position = "right") +
  theme(plot.title = element_text(hjust = 0.5))

ggplot(ipr, aes(x = Subcategory, y = Percent, fill = `Food Security Type`)) + 
  geom_bar(stat = "identity") +
  scale_fill_paletteer_d('ggsci::orange_material') +
  labs(title = "Barplot: Food Security by Income to Poverty Ratio") +
  theme_minimal() +
  theme(legend.position = "right") +
  theme(plot.title = element_text(hjust = 0.5))

ggplot(ar, aes(x = Subcategory, y = Percent, fill = `Food Security Type`)) + 
  geom_bar(stat = "identity") +
  scale_fill_paletteer_d('ggsci::orange_material') +
  labs(title = "Barplot: Food Security by Area of Residence") +
  theme_minimal() +
  theme(legend.position = "right") +
  theme(plot.title = element_text(hjust = 0.5))

ggplot(cgr, aes(x = Subcategory, y = Percent, fill = `Food Security Type`)) + 
  geom_bar(stat = "identity") +
  scale_fill_paletteer_d('ggsci::orange_material') +
  labs(title = "Barplot: Food Security by Census Geographic Region") +
  theme_minimal() +
  theme(legend.position = "right") +
  theme(plot.title = element_text(hjust = 0.5))
```
```{r message = FALSE, warning = FALSE, fig.dim = c(10,6)}
ggplot(ct_trim, aes(x = Category, y = Percent, fill = `Food Security Type`)) + 
  geom_bar(stat = "identity") +
  scale_fill_paletteer_d('ggsci::orange_material') +
  labs(title = "Barplot: Food Security by Child Household") +
  theme_minimal() +
  theme(legend.position = "right") +
  theme(plot.title = element_text(hjust = 0.5)) +
  coord_flip()

ggplot(emp, aes(x = `Sub-subcategory`, y = Percent, fill = `Food Security Type`)) + 
  geom_bar(stat = "identity") +
  scale_fill_paletteer_d('ggsci::orange_material') +
  labs(title = "Barplot: Food Security by Employment") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  theme(plot.title = element_text(hjust = 0.5)) +
  coord_flip()

ggplot(edu, aes(x = `Sub-subcategory`, y = Percent, fill = `Food Security Type`)) + 
  geom_bar(stat = "identity") +
  scale_fill_paletteer_d('ggsci::orange_material') +
  labs(title = "Barplot: Food Security by Education") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  theme(plot.title = element_text(hjust = 0.5)) +
  coord_flip()

ggplot(dis, aes(x = `Sub-subcategory`, y = Percent, fill = `Food Security Type`)) + 
  geom_bar(stat = "identity") +
  scale_fill_paletteer_d('ggsci::orange_material') +
  labs(title = "Barplot: Food Security by Disability Status") +
  theme_minimal() +
  theme(legend.position = "bottom") +
  theme(plot.title = element_text(hjust = 0.5)) +
  coord_flip()
```

## Sources
Food Insecurity Data:

https://www.ers.usda.gov/topics/food-nutrition-assistance/food-security-in-the-us/interactive-charts-and-highlights#States
