---
title: "DATA608 Assignment 6"
author: "Steven Gonzalez"
output: html_document
---

## Prompt
The United Nations Food and Agriculture Organization publication, The State of Food Security and Nutrition in the World 2022 (https://www.fao.org/documents/card/en/c/cc0639en) might lead one to the conclusion that it's an elsewhere problem. That the people who are suffering malnutrition and starvation are "elsewhere", not in our backyard. For this assignment you will need to take a closer look here at home (the US).

## R Packages
```{r message = FALSE, warning = FALSE}
library(tidyverse)
library(httr)
library(jsonlite)
library(dplyr)
library(ggplot2)
library(reshape2)
library(usmap)
library(tinytex)
```

## Data Import
```{r message = FALSE, warning = FALSE}
url <- "https://raw.githubusercontent.com/Stevee-G/Data608/refs/heads/main/Assignment6/foodsecurity_csv_datafiles/foodsecurity-all-households-2024.csv"
url1 <- "https://raw.githubusercontent.com/Stevee-G/Data608/refs/heads/main/Assignment6/foodsecurity_csv_datafiles/foodsecurity-child-trends-2024.csv"
url2 <- "https://raw.githubusercontent.com/Stevee-G/Data608/refs/heads/main/Assignment6/foodsecurity_csv_datafiles/foodsecurity-educ-emp-dis-2024.csv"
url3 <- "https://raw.githubusercontent.com/Stevee-G/Data608/refs/heads/main/Assignment6/foodsecurity_csv_datafiles/foodsecurity-hh-with-children-2024.csv"
url4 <- "https://raw.githubusercontent.com/Stevee-G/Data608/refs/heads/main/Assignment6/foodsecurity_csv_datafiles/foodsecurity-state-2024.csv"

all_hh <- read_csv(url)
ct <- read_csv(url1)
eed <- read_csv(url2)
hh_wc <- read_csv(url3)
state <- read_csv(url4)

glimpse(all_hh)
glimpse(ct)
glimpse(eed)
glimpse(hh_wc)
glimpse(state)
```

## Data Wrangling
```{r message = FALSE, warning = FALSE}
atl_trim <- atlantic_df %>% 
  mutate(year = format(as.Date(atlantic_df$date,
                               format = "%Y%m%d"), "%Y")) %>%
  arrange(desc(max_speed)) %>% 
  distinct(track_id, .keep_all = TRUE) %>% 
  arrange(year, track_id) %>% 
  select(year, track_id, max_speed, status) %>% 
  filter(year >= 1950)
glimpse(atl_trim)

pa_trim <- pacific_df %>% 
  mutate(year = format(as.Date(pacific_df$date,
                               format = "%Y%m%d"), "%Y")) %>%
  arrange(desc(max_speed)) %>% 
  distinct(track_id, .keep_all = TRUE) %>% 
  arrange(year, track_id) %>% 
  select(year, track_id, max_speed, status) %>% 
  filter(year >= 1950)
glimpse(pa_trim)

storms <- bind_rows(atl_trim, pa_trim) %>% 
  mutate(year = as.integer(year),
         max_speed = as.integer(max_speed)) %>% 
  arrange(year, track_id)
glimpse(storms)

temp_trim <- temp_df %>% 
  mutate(temperature = smoothing) %>% 
  filter(year >= 1950) %>% 
  select(year, temperature)
glimpse(temp_trim)

all <- full_join(storms, temp_trim,
                        join_by(year == year),
                        relationship = "many-to-one")
glimpse(all)

all_trim <- all %>% 
  group_by(year,temperature) %>% 
  summarize(
    n_storms = n(),
    n_hurricane = sum(status == "HU"),
    p_hurricane = round(n_hurricane/n_storms,2),
    avg_speed = round(mean(max_speed),2)
  )
glimpse(all_trim)
```

## Data Visualizations
```{r message = FALSE, warning = FALSE}
ggplot(all_trim, aes(x = year, y = n_storms, fill = n_storms)) +
  geom_col() +
  scale_fill_gradient2(low = "whitesmoke", mid = "darkgrey" , high = "darkblue",
                         midpoint=median(all_trim$n_storms)) +
  labs(title = "Bar Plot: Storm Activity Since 1950",
       x = "Year",
       y = "") +
  theme_minimal() +
  theme(legend.position="none",
        plot.title = element_text(hjust = 0.5))

ggplot(all_trim, aes(x = year, y = n_hurricane, fill = n_hurricane)) +
  geom_col() +
  scale_fill_gradient2(low = "violet", mid = "darkviolet" , high = "darkblue",
                         midpoint=median(all_trim$n_hurricane)) +
  labs(title = "Bar Plot: Hurricane Activity",
       x = "Year",
       y = "") +
  theme_minimal() +
  theme(legend.position="none",
        plot.title = element_text(hjust = 0.5))

ggplot(all_trim, aes(x = year, y = p_hurricane, color = p_hurricane)) +
  geom_line(linewidth = 2) +
  scale_colour_gradient2(low = "violet", mid = "darkviolet" , high = "darkblue",
                         midpoint=median(all_trim$p_hurricane)) +
  labs(title = "Line Plot: Proportion of Hurricanes per Storm Season",
       x = "Year",
       y = "Proportion") +
  theme_minimal() +
  theme(legend.position="none",
        plot.title = element_text(hjust = 0.5))

ggplot(all_trim, aes(x = year, y = avg_speed, color = avg_speed)) +
  geom_line(linewidth = 2) +
  scale_colour_gradient2(low = "whitesmoke", mid = "darkgrey" , high = "darkblue",
                         midpoint=median(all_trim$avg_speed)) +
  labs(title = "Line Plot: Average Speed per Storm Season",
       x = "Year",
       y = "Speed") +
  theme_minimal() +
  theme(legend.position="none",
        plot.title = element_text(hjust = 0.5))

ggplot(all_trim, aes(x = year, y = temperature, color = temperature)) +
  geom_line(linewidth = 2) +
  scale_colour_gradient2(low = "yellow", mid = "orange" , high = "red",
                         midpoint=median(all_trim$temperature)) +
  labs(title = "Line Plot: Rise in Global Temperature Over the Years",
       x = "Year",
       y = "Temp (C)") +
  theme_minimal() +
  theme(legend.position="none",
        plot.title = element_text(hjust = 0.5))

ggplot(all_trim, aes(x=temperature, y=n_storms)) +
  geom_point() +
  labs(title = "Scatterplot: Temperature vs Number of Storms",
       x = "Temp (C)",
       y = "Storms") +
  geom_smooth(color = "darkblue") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

ggplot(all_trim, aes(x=temperature, y=p_hurricane)) +
  geom_point() +
  labs(title = "Scatterplot: Temperature vs Proportion of Hurricanes",
       x = "Temp (C)",
       y = "Hurricane Proportion") +
  geom_smooth(color = "darkviolet") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

ggplot(all_trim, aes(x=temperature, y=avg_speed)) +
  geom_point() +
  labs(title = "Scatterplot: Temperature vs Storm Speed",
       x = "Temp (C)",
       y = "Speed") +
  geom_smooth(color = "darkgrey") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

## Sources
Food Insecurity Data:

https://www.ers.usda.gov/topics/food-nutrition-assistance/food-security-in-the-us/interactive-charts-and-highlights#States
